---
title: "Exercise Prediction"
author: "Michael Bristow"
date: "14 December 2017"
output:
  pdf_document: default
  html_document: default
---

#Practical Machine Learning Peer Assessment - Exercide Prediciton

##Executive Summary
This report predicts the manner in which users of exercise devices perform the exercise.

##Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Data Used
The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

Get data files to use for training and testing
```{r}
#check if data dorectory exists
library(caret)
library(randomForest)
set.seed(888)
         
         
if(!file.exists("./data")){
  dir.create("./data")
}
#download files if required
trainingFile <- "./data/pml-training.csv"
traininjgURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"

if (!file.exists(trainingFile)) {
  download.file(traininjgURL, destfile=trainingFile)
}

testingFile <- "./data/pml-testing.csv"
testingURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

if (!file.exists(testingFile)) {
  download.file(testingURL, destfile=testingFile)
}
```

##Data preparation
Load data
```{r}
trainData <- read.csv("./data/pml-training.csv",header=T,sep=",",na.strings=c("NA",""))
testData <- read.csv("./data/pml-testing.csv",header=T,sep=",",na.strings=c("NA",""))
```
Basic data validation
```{r}
#head(trainData)
dim(trainData)
dim(testData)
#nrow(trainData)
```
The training data contains `r nrow(trainData)` rows and we will be preicting on the variable "classe"

##Data cleaning/pre processing
```{r}
#split into training & validating
trainData <- trainData[,-1] # Remove the first column that represents a ID Row
inTrain <- createDataPartition(y=trainData$classe, p=0.7, list=FALSE)
trainingData <- trainData[inTrain,]
validationData <- trainData[-inTrain,]


sum((colSums(!is.na(trainingData[,-ncol(trainingData)])) < 0.6*nrow(trainingData)))
# Number of cols with less than 50% of data
Keep <- c((colSums(!is.na(trainingData[,-ncol(trainingData)])) >= 0.6*nrow(trainingData)))
trainingData   <-  trainingData[,Keep]
validationData   <-  validationData[,Keep]


```

##Data Modelling
We will be using random forest method

```{r cache=TRUE}
fmodel <- randomForest(classe~.,data=trainingData)
fmodel
```

##Data Evaluation
Variable importance of the model and also produce confusion matrix
```{r cache=TRUE}
importance(fmodel)

confusionMatrix(predict(fmodel,newdata=validationData[,-ncol(validationData)]),validationData$classe)


aCC <-c(as.numeric(predict(fmodel,newdata=validationData[,-ncol(validationData)])==validationData$classe))

aCC <-sum(aCC)*100/nrow(validationData)

```
Model Accuracy as tested over Validation set = `r aCC`%

##Testing the model
We mow test the model on the testing data and print out the predictions

Data must be cleaned in the same way as training data o ensure same datat format
```{r}
testData <- testData[,-1]
testData <- testData[,Keep]
testData <- testData[,-ncol(testData)]
testing <- rbind(trainingData[100, -59] , testData) 

predictions <- predict(fmodel,newdata=testing)
predictions
```


